Box 1 - TDT Curves

### 2023-09-21 ms/ms.qmd
Updated documents

@@

rerun = FALSE
if(rerun){
    prior = list(R = list(V = diag(3), nu=1.002), 
               G = list(G1=list(V = diag(3), nu = 1.002)))

@@

  mcmcglmm_model <- MCMCglmm::MCMCglmm(cbind(ctmax, cfts, tdt_z) ~ -1 + trait, random = ~us(trait):animal, rcov = ~us(trait):units ,pedigree=tree3,burnin=1000, nitt = 300000, thin = 5, family = rep("gaussian", 3), prior = prior, pl = TRUE, pr = TRUE, data = data)

@@

  mcmcglmm_model_z_trans <- MCMCglmm::MCMCglmm(cbind(z_ctmax, z_cfts, z_tdt_z) ~ -1 + trait, random = ~us(trait):animal, rcov = ~us(trait):units, pedigree=tree3,burnin=1000, nitt = 300000, thin = 5, family = rep("gaussian", 3), prior = prior, pl = TRUE, pr = TRUE, data = data)

  # Ditch phylogeny to see if there is support for including it or not. DIC is WAY lower for phylogenetic model, so good to include it.
  prior2 = list(R = list(V = diag(3), nu=1.002))
  mcmcglmm_model_z_trans_nophylo <- MCMCglmm::MCMCglmm(cbind(z_ctmax, z_cfts, z_tdt_z) ~ -1 + trait, rcov = ~us(trait):units, pedigree=tree3,burnin=1000, nitt = 300000, thin = 5, family = rep("gaussian", 3), prior = prior2, pl = TRUE, pr = TRUE, data = data)

@@

# Extract the estimated / imputed Z for each species, but drop internal ancestral nodes
  z_blups <- sol[, grep("traittdt_z.animal", colnames(sol))]
  z_blups <- z_blups[,-grep("Node", colnames(z_blups))]
  head(z_blups)

# Remember that these are BLUPs. They are deviations from the intercept. So, to get back to the z we need to add the intercept to each of them.
  z <- z_blups + sol[, "traittdt_z"]

# Clean up names so they are just species names
  colnames(z) <- gsub("traittdt_z.animal.", "", colnames(z))

# Have a look at how well the model gets the actual data correct.
  z_known <- data %>% filter(!is.na(tdt_z)) %>% dplyr::select(species, tdt_z) %>% distinct()
    z_est <- apply(z, 2, function(x) median(x))
    z_est <- z_est[which(colnames(z) %in% z_known$species)]
    z_est <- data.frame(z_est)
    z_est$species <- row.names(z_est)
  check_z <- left_join(z_known, z_est, by = "species")

  # Check phylogeneic signal. Need to think about the covariance here. 
    # CTMax - Yes
      phylo_ctmax <- vcv[,"traitctmax:traitctmax.animal"]
      units_ctmax <- vcv[,"traitctmax:traitctmax.units"]

      h2_ctmax = phylo_ctmax / (phylo_ctmax + units_ctmax)
      posterior.mode(h2_ctmax)
      HPDinterval(h2_ctmax)

    # CFL - Yes
      phylo_cfls <- vcv[,"traitcfts:traitcfts.animal"]
      units_cfls <- vcv[,"traitcfts:traitcfts.units"]

      h2_cfls = phylo_cfls / (phylo_cfls + units_cfls)
      posterior.mode(h2_cfls)
      HPDinterval(h2_cfls)

    # z - Yes
      phylo_z <- vcv[,"traittdt_z:traittdt_z.animal"]
      units_z <- vcv[,"traittdt_z:traittdt_z.units"]

      h2_z = phylo_z / (phylo_z + units_z)
      posterior.mode(h2_z)
      HPDinterval(h2_z)
```

# Plot the tree
  plot(tree, cex = 0.2)
```

@@

Powerful multivariate hierachical models (i.e., multi-response models) are capable of dealing with missing data, and provide a means by which multiple physiological thermal tolerance measures can be modelled simultaneously [@Noble2021; @Pottier2023]. When used in conjunction with an understanding of the phylogenetic relationships among species and/or populations, such modelling approaches can provide informed estimates for species missing data and explicitly estimate the covariance between tolerance measures while accounting for their uncertainty. Such insight that can be used to inform policy and conservation-related decision making for species with little to no data.

@@

We integrated static lethal limits ($CT_{max}$) collected by @Heerwaarden2021 and XX with dynamic lethal limit measures from TDT curces [i.e., slope ($z^{TDT}$) and intercept ($CT_{max}^{TDT}$)] estimated by XX. We also collected data to estimate TDT curve parameters for eight additonal species following protocols outlined in @Orsted2022. Critical fertility limits for *Drosophilids* were compiled from @Parratt2021 and @Heerwaarden2021. These data were integrated with data on lethal limits at the species-level. When the same species were sampled multiple times we retained within species variation in models. We then fit the following model in R (vers. `r paste0(R.Version()$major, ".", R.Version()$minor)`)[@R2023] with the package *MCMCglmm* (vers. `r utils::packageVersion("MCMCglmm")`) [@Hadfield2010]:

$$
\begin{bmatrix}
 CT_{max} \\
 CT_{max}^{TDT} \\
 z^{TDT} \\
 CFLs
\end{bmatrix} = u_{i,k} + p_{i,k} + sp_{i,k} + e_{i,k}
$$

>>>>

# Plot the tree
  plot(tree, cex = 0.2, type = "radial")
```

@@

rerun = FALSE
if(rerun){
    prior = list(R = list(V = diag(3), nu=1.002), 
               G = list(G1=list(V = diag(3), nu = 1.002),
                        G2=list(V = diag(3), nu = 1.002)))

@@

  mcmcglmm_model <- MCMCglmm::MCMCglmm(cbind(ctmax, cfts, tdt_z) ~ -1 + trait, random = ~us(trait):animal + us(trait):species, rcov = ~us(trait):units ,pedigree=tree3,burnin=1000, nitt = 300000, thin = 5, family = rep("gaussian", 3), prior = prior, pl = TRUE, pr = TRUE, data = data)

@@

  mcmcglmm_model_z_trans <- MCMCglmm::MCMCglmm(cbind(z_ctmax, z_cfts, z_tdt_z) ~ -1 + trait, random = ~us(trait):animal + us(trait):species, rcov = ~us(trait):units, pedigree=tree3,burnin=1000, nitt = 300000, thin = 5, family = rep("gaussian", 3), prior = prior, pl = TRUE, pr = TRUE, data = data)

  # Ditch phylogeny to see if there is support for including it or not. DIC is WAY lower for phylogenetic model, so good to include it.
  prior2 = list(R = list(V = diag(3), nu=1.002), 
               G = list(G1=list(V = diag(3), nu = 1.002)))
  mcmcglmm_model_z_trans_nophylo <- MCMCglmm::MCMCglmm(cbind(z_ctmax, z_cfts, z_tdt_z) ~ -1 + trait, random = ~us(trait):species, rcov = ~us(trait):units, pedigree=tree3,burnin=1000, nitt = 300000, thin = 5, family = rep("gaussian", 3), prior = prior2, pl = TRUE, pr = TRUE, data = data)

@@

# Extract the estimated / imputed Z for each species, but drop internal ancestral nodes
    z_BLUPS <- extract_blups(sol, trait ="tdt_z")
ctmax_blups <- extract_blups(sol, trait ="ctmax")
  cfts_blups <- extract_blups(sol, trait ="cfts")

# Remember that these are BLUPs. They are deviations from the intercept. So, to get back to the z we need to add the intercept to each of them.
      z <- z_BLUPS[["phy"]] + z_BLUPS[["spp"]] + sol[, "traittdt_z"]
  ctmax <- ctmax_blups[["phy"]] + ctmax_blups[["spp"]]+ sol[, "traitctmax"]
   cfts <- cfts_blups[["phy"]] + cfts_blups[["spp"]]+ sol[, "traitcfts"]

# Clean up names so they are just species names
      colnames(z) <- gsub("traittdt_z.animal.", "", colnames(z))
  colnames(ctmax) <- gsub("traitctmax.animal.", "", colnames(ctmax))
   colnames(cfts) <- gsub("traitcfts.animal.", "", colnames(cfts))

# Have a look at how well the model gets the actual data correct.
    check_z <- predict_accuracy(data, z, trait = tdt_z)
check_ctmax <- predict_accuracy(data, ctmax, trait = ctmax)
 check_cfls <- predict_accuracy(data, cfts, trait = cfts) # Something wrong here.

  # Plot the data
    plot(check_z$tdt_z ~ check_z$est, ylab = "True Slope (z)", xlab = "Estimated Slope", pch = 16, cex = 2)
    abline(a = 0, b = 1, col = "red")
    plot(check_ctmax$ctmax ~ check_ctmax$est, ylab = "True CTmax", xlab = "Estimated CTmax", pch = 16, cex = 2)
    abline(a = 0, b = 1, col = "red")
    plot(check_cfls$cfts ~ check_cfls$est, ylab = "True CFLs", xlab = "Estimated CFLs", pch = 16, cex = 2)
    abline(a = 0, b = 1, col = "red")

  # Check phylogeneic signal. Need to think about the covariance here. 
    # CTMax - Yes
      phylo_ctmax <- vcv[,"traitctmax:traitctmax.animal"]
        spp_ctmax <- vcv[,"traitctmax:traitctmax.species"]
      units_ctmax <- vcv[,"traitctmax:traitctmax.units"]

      h2_ctmax = phylo_ctmax / (phylo_ctmax + spp_ctmax + units_ctmax)
      posterior.mode(h2_ctmax)
      HPDinterval(h2_ctmax)

    # CFL - Yes
      phylo_cfls <- vcv[,"traitcfts:traitcfts.animal"]
        spp_cfls <- vcv[,"traitcfts:traitcfts.species"]
      units_cfls <- vcv[,"traitcfts:traitcfts.units"]

      h2_cfls = phylo_cfls / (phylo_cfls + spp_cfls + units_cfls)
      posterior.mode(h2_cfls)
      HPDinterval(h2_cfls)

    # z - Yes
      phylo_z <- vcv[,"traittdt_z:traittdt_z.animal"]
        spp_z <- vcv[,"traittdt_z:traittdt_z.species"]
      units_z <- vcv[,"traittdt_z:traittdt_z.units"]

      h2_z = phylo_z / (phylo_z + spp_z + units_z)
      posterior.mode(h2_z)
      HPDinterval(h2_z)
```

@@

Powerful multivariate hierachical models (i.e., multi-response models) are capable of dealing with missing data, and provide a means by which multiple physiological thermal tolerance measures can be modelled simultaneously [@Noble2021; @Pottier2023]. When used in conjunction with an understanding of the phylogenetic relationships among species and/or populations, such modelling approaches can provide informed estimates for species missing data and explicitly estimate the covariance between tolerance measures while accounting for their uncertainty. Such insight can be used to inform policy and conservation-related decision making for species with little to no data.

@@

We integrated static lethal limits ($CT_{max}$) collected by @Heerwaarden2021 and XX with dynamic lethal limit measures from TDT curces [i.e., slope ($z^{TDT}$) and intercept ($CT_{max}^{TDT}$)] estimated by XX. We also collected data to estimate TDT curve parameters for eight additonal species following protocols outlined in @Orsted2022. Critical fertility limits for *Drosophilids* were compiled from @Parratt2021 and @Heerwaarden2021. These data were integrated with data on lethal limits at the species-level. When the same species were sampled multiple times we retained within species variation in models. We then fit the following model in R (vers. `r paste0(R.Version()$major, ".", R.Version()$minor)`)[@R2023] with the package *MCMCglmm* (vers. `r utils::packageVersion("MCMCglmm")`), which uses data augmentation during model fittng to estimate missing data [@Hadfield2010]:

$$
\begin{bmatrix}
 CT_{max} \\
 z^{TDT} \\
 CFLs
\end{bmatrix} = u_{i,k} + p_{i,k} + sp_{i,k} + e_{i,k}
$$

Models including phylogeny were better supported than models without phylogeny ($\Delta DIC$ = `r mcmcglmm_model_z_trans_nophylo$DIC - mcmcglmm_model_z_trans$DIC`), with all traits exhibiting a moderate to strong phylogenetic signal (Phylogenetic $H^2$: $CT_{max}$ – `r posterior.mode(h2_ctmax)`, 95% CI: `r HPDinterval(h2_ctmax)[,"lower"]` to `r HPDinterval(h2_ctmax)[,"upper"]`; $z^{TDT}$ – `r posterior.mode(h2_z)`, 95% CI: `r HPDinterval(h2_z)[,"lower"]` to `r HPDinterval(h2_z)[,"upper"]`; CFLs – `r posterior.mode(h2_cfls)`, 95% CI: `r HPDinterval(h2_cfls)[,"lower"]` to `r HPDinterval(h2_cfls)[,"upper"]`). The model predicted known data reasonably well for most traits (correlation between actual and estimated data: $z^{TDT}$ – `r cor.test(check_z$tdt_z, check_z$est)$estimate`, 95% CI: `r cor.test(check_z$tdt_z, check_z$est)$conf.int[1]` to `r cor.test(check_z$tdt_z, check_z$est)$conf.int[2]`, p < 0.001).

```{r, fig-treesub}
#| label: fig-treesub
#| fig-cap: Species included in the dataset and their phylogenetic relationships.

# prune tree
t <- tree_checks(data, tree, dataCol = "species", type = "prune")

layout.matrix <- matrix(c(1, 1, 1, 2,3,4), nrow = 3, ncol = 2) # Phylogeny takes up 3 panels (plot = 1) and the others take up a single panel.

layout(mat = layout.matrix) 
# Plot 1
plot(t, type = "tidy", cex = 0.65)

# Plot 2
plot(check_z$tdt_z ~ check_z$est, ylab = "TDT Slope (z)", xlab = "Estimated TDT Slope", pch = 16, cex = 3, cex.lab = 2)
abline(a = 0, b = 1, col = "red")
